### 2023-04-08 main/chapter_experiment.tex
type some lhc notes from reading
begin lhc accelerator complex words

\section{The Large Hadron Collider}
\label{sec:atlas_lhc}

% the LHC
% where the protons come from
Figure~\ref{fig:experiment_cern_complex}

@@

% the accelerator complex
% RF cavities
% magnets (dipole, quadrupole etc)
% size of beam
% luminosity and its uncertainty
% pile-up
\TODO{Distribution in the mean number of primary vertices per crossing, where the
variability in that mean arises from differing beam conditions over run-time.}
Figure~\ref{fig:experiment_run_2_mu}

\begin{figure}[tp]
\centering
\includegraphics[width=0.62\textwidth]{figures/atlas_pileup_mu_2015_2018.pdf}
\caption[
Mean number interactions per bunch crossing in Run~2 of the LHC
]{%
Mean number interactions per bunch crossing in Run~2 of the LHC;
rates presented are weighted by luminosity and sub-divided by year of
operation.
This figure includes all data that \atlas\ recorded with stable LHC proton
beams.
The reported integrated luminosities therefore exceed those we present later
in Table~\ref{tab:experiment_run2_state}, which are further filtered to require
other data-goodness criteria.
This figure is reproduced from~\cite{atlas_public_luminosity}.
}
\label{fig:experiment_run_2_mu}
\end{figure}

>>>>

\section{The Large Hadron Collider at CERN}
\label{sec:atlas_lhc}
% The LHC
% largest piece
The Large Hadron Collider (LHC) is the largest and most powerful piece of
CERN's extensive particle accelerator complex.
It lives in a $26.7\,\mathrm{km}$ (circumference) tunnel, which is buried
around $100\,\mathrm{m}$ under France and Switzerland at CERN, Geneva.
It and comprises two rings that contain and accelerate counter-rotating bunches
of particles.
These accelerated bunches are brought into near-head-on collisions at
the four large detector experiments:
\atlas~\cite{atlas2008experiment},
\lhcb~\cite{lhcb2008experiment},
\cms~\cite{cms2008experiment}, and
\alice~\cite{alice2008experiment}.
Although proton-proton collisions are the LHC's primary subject, it is also
operated for heavy ion collisions including lead ($^{208}\mathrm{Pb}^{82+}$)
and oxygen ($^{16}\mathrm{O}^{4+}$)~\cite{
Brewer:2021kiv,
lhc2008machine,
lhcDesignreport2004
}.
Heavy ion collisions are primarily studied by the \alice\ experiments, but are
also analysed by the general-purpose detectors \atlas\ and \cms.
Since proton collision data are the topic of this thesis, this section focuses
on proton operations.

% does not stand alone
The LHC does not stand alone.
It implements the final boosts and collisions of particles, but requires
earlier stages of CERN's systems to inject the protons at suitably high
energies for it to continue their acceleration.
In brief, the accelerator chain flows as follows:
% Accelerator complex
\begin{enumerate}
% TODO where the protons come from
\item The protons come from
\end{enumerate}
This sequence is a subset of CERN's full accelerator complex,
which is illustrated in
Figure~\ref{fig:experiment_cern_complex}~\cite{Haffner:1621894}.

@@

% the accelerator complex
% - sequence of contributing accelerators and their beam energies
% beampipe restricted by size of lep tunnel
% - required particle count makes antiprotons hard
% - proton-proton same charge requires separate beampipes
% - size means most cost efficient to put both beams in one cryostat
% - beampipes therefore mechanically and magnetically coupled
% - superconducting magnets at around 2 to 4 Kelvin
% - cooled with superfluid helium
% - low heat capacity, danger of heating from beam radiation
% Radiofrequency cavities
% - radio waves at 400 MHz such that slow protons accelerated, fast decelerated
% - in LHC, accelerates from 450 GeV to 6.5 TeV (7 target)
% magnets
% - dipole for steering
% - quadrupole for focussing (pairs)
% - higher orders why? septupole at beam entry
% TODO size of beam
% TODO luminosity and its uncertainty
% TODO pile-up
\TODO{Distribution in the mean number of primary vertices per crossing, where the
variability in that mean arises from differing beam conditions over run-time.}
Figure~\ref{fig:experiment_run_2_mu}

\begin{figure}[tp]
\centering
\includegraphics[width=0.7\textwidth]{figures/atlas_pileup_mu_2015_2018.pdf}
\caption[
Mean number interactions per bunch crossing in Run~2 of the LHC
]{%
Mean number interactions per bunch crossing in Run~2 of the LHC;
rates presented are weighted by luminosity and sub-divided by year of
operation.
This figure includes all data that \atlas\ recorded with stable LHC proton
beams.
The reported integrated luminosities therefore exceed those we present later
in Table~\ref{tab:experiment_run2_state}, which are further filtered to require
other data-goodness criteria.
This figure is reproduced from~\cite{atlas_public_luminosity}.
}
\label{fig:experiment_run_2_mu}
\end{figure}
